---
title: "STAT600 HW1"
author: "Muyang Shi"
date: "2024-01-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
library(SimpLin)
library(dplyr)
library(knitr)
library(ggplot2)
library(latex2exp)
library(gridExtra)
library(foreach)
library(doParallel)
numcores <- detectCores()
registerDoParallel(numcores - 2)
```

## Problem 2

To install my package, please download the `SimpLin_1.0.tar.gz` file, github link here: https://github.com/muyangshi/STAT600Homework/tree/main/HW1.

## Problem 3

### Simulate Dataset

```{r}
set.seed(1)
Dataset100 <- foreach(i = 1:100) %do% {
  x <- rnorm(100)
  e <- rnorm(100)
  y <- 1 - x + e
  list(x = x, y = y)
}
```

### Base R `lm()`

```{r}
# Base lm()
lm_base <- foreach(i = 1:100) %dopar% {
  x <- Dataset100[[i]]$x
  y <- Dataset100[[i]]$y
  elapsed <- system.time({mod.base <- lm(y~x)})[3]
  coef <- c(mod.base$coefficients)
  beta0_ci <- confint(mod.base)[1,]
  beta1_ci <- confint(mod.base)[2,]
  yhat <- mod.base$fitted.values
  resids <- mod.base$residuals
  list(time=elapsed,coef=coef,beta0_ci=beta0_ci,beta1_ci=beta1_ci,yhat=yhat,resids=resids)
}
stopImplicitCluster()
```

### SimpLinCpp

```{r}
lm_cpp <- foreach(i = 1:100) %dopar% {
  x <- Dataset100[[i]]$x
  y <- Dataset100[[i]]$y
  elapsed <- system.time({mod.cpp <- SimpLinR(x,y)})[3]
  coef <- c(mod.cpp$Coefficients)
  beta0_ci <- mod.cpp$`95% CI`[1,]
  beta1_ci <- mod.cpp$`95% CI`[2,]
  yhat <- mod.cpp$`Predicted Values`
  resids <- mod.cpp$Residuals
  list(time=elapsed,coef=coef,beta0_ci=beta0_ci,beta1_ci=beta1_ci,yhat=yhat,resids=resids)
}
stopImplicitCluster()
```

## Problem 4

Both the base R `lm()` and the Cpp `SimpLin` produce same result (in terms of bias, coverage probability, mean squared error for the regression coefficients as well as predictive mse for $\hat{y}$) like they should be. 
Time-wise, the cpp is faster than the base R `lm()`.

A table comparing the two methods' performance is included below, as well as histogram of the estimated coefficients.

```{r, echo = FALSE}
time.base <- mean(sapply(lm_base,FUN=function(mod){mod$time}))
time.cpp  <- mean(sapply(lm_cpp,FUN=function(mod){mod$time}))

bias.beta0.base <- mean(sapply(lm_base,FUN=function(mod){mod$coef[1] - 1}))
bias.beta0.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){mod$coef[1] - 1}))

bias.beta1.base <- mean(sapply(lm_base,FUN=function(mod){mod$coef[2] - (-1)}))
bias.beta1.cpp  <- mean(sapply(lm_cpp ,FUN=function(mod){mod$coef[2] - (-1)}))

cover.beta0.base <- mean(sapply(lm_base,FUN=function(mod){mod$beta0_ci[1] <= 1 & 1 <= mod$beta0_ci[2]}))
cover.beta0.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){mod$beta0_ci[1] <= 1 & 1 <= mod$beta0_ci[2]}))

cover.beta1.base <- mean(sapply(lm_base,FUN=function(mod){mod$beta1_ci[1] <= -1 & -1 <= mod$beta1_ci[2]}))
cover.beta1.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){mod$beta1_ci[1] <= -1 & -1 <= mod$beta1_ci[2]}))

mse.beta0.base <- mean(sapply(lm_base,FUN=function(mod){mod$coef[1] - 1})^2)
mse.beta0.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){mod$coef[1] - 1})^2)

mse.beta1.base <- mean(sapply(lm_base,FUN=function(mod){mod$coef[2] - (-1)})^2)
mse.beta1.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){mod$coef[2] - (-1)})^2)

mse.pred.base <- mean(sapply(lm_base,FUN=function(mod){(mod$resids)^2}))
mse.pred.cpp  <- mean(sapply(lm_cpp, FUN=function(mod){(mod$resids)^2}))

rbind(c(time.base, time.cpp), 
      c(bias.beta0.base,bias.beta0.cpp),
      c(bias.beta1.base,bias.beta1.cpp), 
      c(cover.beta0.base,cover.beta0.cpp),
      c(cover.beta1.base,cover.beta1.cpp), 
      c(mse.beta0.base,mse.beta0.cpp), 
      c(mse.beta1.base,mse.beta1.cpp), 
      c(mse.pred.base,mse.pred.cpp)) %>%
  data.frame() %>%
  `rownames<-`(c("Avg Time (secs)", 
                 "Bias: b0", 
                 "Bias: b1",
                 "Empirical Coverage: b0", 
                 "Empirical Coverage: b1",
                 "MSE: b0", 
                 "MSE: b1", 
                 "Prediction MSE")) %>%
  kable(col.names = c("Base R", "Rcpp"),
        caption = "Performace Comparison between Base R and Rcpp")
```

```{r, echo = FALSE, out.height='40%', fig.align='center'}
hist.b0.base <- ggplot(data.frame(beta0=sapply(lm_base,FUN=function(mod){mod$coef[1]}))) +
  geom_histogram(aes(x=beta0, y = after_stat(density)),bins=10) +
  theme_bw() +
  labs(x = TeX("$\\hat{\\beta}_{0}$"), 
       y = 'density',
       title = 'Base R Beta0 Estimates')
hist.b0.cpp <- ggplot(data.frame(beta0=sapply(lm_cpp,FUN=function(mod){mod$coef[1]}))) +
  geom_histogram(aes(x=beta0, y = after_stat(density)),bins=10) +
  theme_bw() +
  labs(x = TeX("$\\hat{\\beta}_{0}$"), 
       y = 'density',
       title = 'Rcpp Beta0 Estimates')
grid.arrange(hist.b0.base,hist.b0.cpp, nrow=1)

hist.b1.base <- ggplot(data.frame(beta1=sapply(lm_base,FUN=function(mod){mod$coef[2]}))) +
  geom_histogram(aes(x=beta1, y = after_stat(density)),bins=10) +
  theme_bw() +
  labs(x = TeX("$\\hat{\\beta}_{1}$"), 
       y = 'density',
       title = 'Base R Beta1 Estimates')
hist.b1.cpp <- ggplot(data.frame(beta1=sapply(lm_cpp,FUN=function(mod){mod$coef[2]}))) +
  geom_histogram(aes(x=beta1, y = after_stat(density)),bins=10) +
  theme_bw() +
  labs(x = TeX("$\\hat{\\beta}_{1}$"), 
       y = 'density',
       title = 'Rcpp Beta1 Estimates')
grid.arrange(hist.b1.base,hist.b1.cpp, nrow=1)

```