---
title: "SimpLin_Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SimpLin_Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SimpLin)
```

The R function, `SimpLinR` is exposed to the user.
It takes in numeric vectors $\boldsymbol{x}$ and $\boldsymbol{y}$, outputs estimated regression coefficients $\hat{\beta_0}$ and $\hat{\beta_1}$ and their corresponding standard errors and 95% confidence intervals, residuals, and predicted values. 

Example:
Assuming the data is generated under $$y = x + \epsilon$$ where $$\epsilon \sim N(0,1)$$
```{r}
x <- 1:30
y <- x + rnorm(30)
slr <- SimpLinR(x,y)
```

The estimated regression coefficients are, respectively:
```{r}
slr$Coefficients
```

Their corresponding standdard errors are, respectively:
```{r}
slr$`Std. Errors`
```

The 95% confidence intervals are, respectively (with the first row representing $\hat{\beta_0}$ and the second row representing $\hat{\beta_1}$):
```{r}
slr$`95% CI`
```

The residuals can be accessed through:
```{r}
slr$Residuals
```

The predicted values can be accessed with:
```{r}
slr$`Predicted Values`
```